---
title: "Test_RMkdwn"
author: "TK51"
date: "2023-04-08"
output: html_document
---

## Markdown syntax

You can use regular markdown rules in your R Markdown document. Once you knit your document, the output will display text formatted according to the following simple rules.

## Formatting Text

Here are a few common formatting commands:

*Italic*

Italic

**Bold**

Bold

This is `code` in text

This is code in text

# Header 1

Header 1

## Header 2

Header 2

Note that when a # symbol is placed inside a code chunk it acts as a normal R comment, but when placed in text it controls the header size.

* Unordered list item

Unordered list item

1. Ordered list item

Ordered list item

[Link](https://www.google.com)

Link

$A = \pi \times r^{2}$

Rendered equation example

The $ symbols tells R markdown to use LaTeX equation syntax.

```{r}
norm <- rnorm(100, mean = 0, sd = 1)
```

```{r}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
```

```{r, echo = FALSE}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
```

```{r}
library(dplyr)
```

```{r}
print("yey!")
```
```{r, fig.width = 4, fig.height = 3.5}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
boxplot(B~A,data=dataframe)
```

```{r}
dataframe
```
```{r}
library(knitr)
kable(dataframe, digits = 2)
```







