---
title: "Test_RMkdwn"
author: "TK51"
date: "2023-04-08"
output:
  pdf_document:
  html_document: 
geometry: left = 0.5cm, right = 1cm, top = 1cm, bottom = 1cm
---

# YES, IT IS! The starter (2nd timeR)!

## Markdown syntax

You can use regular markdown rules in your R Markdown document. Once you knit your document, the output will display text formatted according to the following simple rules.

## Formatting Text

Here are a few common formatting commands:

*Italic*

Italic

**Bold**

Bold

This is `code` in text

This is code in text

# Header 1

Header 1

## Header 2

Header 2

Note that when a # symbol is placed inside a code chunk it acts as a normal R comment, but when placed in text it controls the header size.

* Unordered list item

Unordered list item

1. Ordered list item

Ordered list item

[Link](https://www.google.com)

Link

$A = \pi \times r^{2}$

Rendered equation example

The $ symbols tells R markdown to use LaTeX equation syntax.

```{r}
norm <- rnorm(100, mean = 0, sd = 1)
```

```{r}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
```

```{r, echo = FALSE}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
```

```{r}
library(dplyr)
```

```{r}
print("yey!")
```
```{r, fig.width = 4, fig.height = 3.5}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
dataframe <- data.frame(A, B)
print(dataframe)
boxplot(B~A,data=dataframe)
```

```{r}
dataframe
```
```{r}
library(knitr)
kable(dataframe, digits = 2)
```

```{r}
library(pander)
plant <- c("a", "b", "c")
temperature <- c(20, 20, 20)
growth <- c(0.65, 0.95, 0.15)
dataframe <- data.frame(plant, temperature, growth)
emphasize.italics.cols(3)   # Make the 3rd column italics
pander(dataframe)           # Create the table
```
| Plant | Temp. | Growth |
|:------|:-----:|-------:|
| A     | 20    | 0.65   |
| B     | 20    | 0.95   |
| C     | 20    | 0.15   |

\pagebreak
```{r warning=FALSE}
library(broom)
library(pander)
A <- c(20, 15, 10)
B <- c(1, 2, 3)

lm_test <- lm(A ~ B)            # Creating linear model

table_obj <- tidy(lm_test)      # Using tidy() to create a new R object called table

pander(table_obj, digits = 3)   # Using pander() to view the created table, with 3 sig figs 
```

\pagebreak
```{r echo=FALSE}
library(dplyr)
library(pander)
library(knitr)
getwd()
```
```{r}
edidiv <- read.csv("edidiv.csv")
richness <-
  edidiv %>%
  group_by(taxonGroup) %>%
  summarise(Species_richness = n_distinct(taxonName))

richness

pander(richness)

kable(richness)
```










