---
title: "RMarkdown_Tutorial_TK"
author: "TK51"
date: "2023-04-09"
output:
  pdf_document: default
  html_document: default
geometry: left = 0.5cm, right = 1cm, top = 1cm, bottom = 1cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# The Task

Use this example R script to practice compiling an R Markdown file, using the 
tutorial materials provided at: [ourcodingclub.github.io](ourcodingclub.github.io/2016/11/24/rmarkdown-1.html)

Follow through the tutorial to make a well commented, easy to follow record of
what is going on so that others can easily follow.

## Preabmle

### Loading packages
```{r, message=FALSE, warning=FALSE}
library(dplyr)   # for data manipulation
library(pander)  # to create pretty tables
```

### Loading biodiversity data
```{r, include=FALSE}
setwd("/home/t51/DataAnalytics/R-Playground/coding_club")
edidiv <- read.csv("datasets/edidiv.csv")
```

## Data exploration

This data is a publicly available dataset of occurrence records for many animal, 
plant, and fungi species, for 2000-2016 from the NBN Gateway

### What is the species richness across taxonomic groups?

Constructing a table of species richness in each taxonomic group

```{r, results='asis'}
richness <- 
  edidiv %>%
    group_by(taxonGroup) %>%
    summarise(Species_richness = n_distinct(taxonName))

pander(richness)
```

Creating a barplot of species richness in each taxonomic group
```{r, fig.align="center", fig.width=15, fig.height=8}
barplot(richness$Species_richness, 
        names.arg = richness$taxonGroup, 
        xlab = "Taxa", ylab = "Number of species", 
        ylim = c(0,600)
        ) 
```
### Determining what the most common species is in each taxonomic group 
```{r}
# Create a vector of most abundant species per taxa
max_abund <-
  edidiv %>%
    group_by(taxonGroup) %>%
    summarise(taxonName = names(which.max(table(taxonName))))

max_abund
```

```{r}
# Joining the two data frames together, using "taxonGroup" as the reference
richness_abund <- inner_join(richness, max_abund, by = "taxonGroup")
```

```{r}
# Renaming the headers of the tables, and viewing the data frame
richness_abund <- rename(richness_abund, Most_abundant =  taxonName, Taxon = taxonGroup)

richness_abund

pander(richness_abund)
```
```{r}
library(knitr)
kable(richness_abund)
```
\pagebreak
### Things to think about:

  - Which bits of code need to be displayed in the final .html file?
  - How can the formatting of the R markdown file be improved?

Experiment with other demo R scripts in the repo, or your own scripts for further practice! 

  - RMarkdown_Demo_1.R 
  - RMarkdown_Demo_2.R 
  - RMarkdown_Demo_3.R
